Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-08-12 anzjs2018/response.txt

Thank you for all of the very constructive feedback, I hope I've covered them adequately. Below I've replied to each point following the order/layout of the original comments.

Referee 1:
The issue of constant stopping probabilities and dwell times is something we're planning to explore with historical data, which I've now mentioned in the discussion (lines 331-333). After the model specification, I have now addressed the likely consequences of making the assumption (lines 187-191).

- I've clarified dimensions of the state vector (lines 131-133).
- The measurement function was a function of X and a sampled value of noise for generality, but I've changed this to h(x) + e in (1) to avoid confusion.
- I've re-expressed the likelihood equations to more clearly describe the relationship between X and Y, through the functions g and h (section 3.1.2)
- Degeneration has been described in more detail (para starting line 158), and PF described through model formulae instead of as an algorithm (section 3.1.1); included specification of N_eff value we used (line 170).
- Continuity improved: "intuitive likelihood" is followed with reasoning (line 153)

- spelling fixed, now using Figure/Section etc consistently, and using all UK English.
- description of network construction (removing route dependencies) has been clarified (section 2.1)
- figures improved to be consistent and clearer
- added citations for R, dplyr, and ggplot2 (all used to summarise and present simulation results)

Referee 2:
1. I've added a reference to the Ulmke 2006 paper (para starting line 35), which demonstrates the advantage of the PF when there's sparse sampling.

2. The main advantage of our approach is that we don't need to snap points to the route. I've included an example where "snapping" can fall over in previous (i.e., KF) approaches do highlight the advantage of avoiding it (bottom of page 5).

3. I've removed z from the vehicle state X and explained how we take advantage of the PF to estimate these values in real-time using eq (4).

4. Added state equations to show that Z is simply a measurement of theta, which is obtained from the PF (top of page 12)

5. Added missing units to parameters.

6. We're in the process of developing a simulation to test travel time estimation as part of the arrival time prediction component of the project, so these aren't available for the current paper.

7. Clarified that no correlation structure is used (line 230).

8. All figures improved/restructured to make them much easier to read and remove any ambiguity.

AE:
I've respecified the PF state equations using Dirac, which I agree is much better, thank you.

Instead of writing out the model algorithm, I've described the model using the suggested notation and briefly mentioned how this is implemented using an iterative algorithm. I don't think there was any advantage to including the full algorithm description, so I decided to remove it.

I've re-written the likelihood section to clarify g, h, and the relationship between X and Y, as well as fixed the definition to keep epsilon on the RHS.

Fixed typos in equation subscript (page 12, eq 3), and the noise term (page 12, eq 1).

And moved future work comments to the end.

### 2019-08-12 anzjs2018/sections/03_models.tex
minor edits

The particle filter provides us with a flexible method
with which to model bus behaviour.
Currently, this model includes
\begin{enumerate}
\item non-constant speed along roads (between observations), and
\item bus stop behaviour, which involves optional stopping and waiting times
    while passengers board and disembark.
\end{enumerate}
The transition function consists of two models,
one for each of the above behaviours.
The first models the speed and distance travelled by the vehicle,
\begin{equation}
\label{eq:trans_newton}
\bX_k^{(i)} = f(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) = 
    \begin{bmatrix}
        x_{k-1}^{(i)} + \delta_k(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
        \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
    \end{bmatrix},
\end{equation}
where $\delta_k = t_k - t_{k-1}$
and $\epsilon_k^{(i)}\sim\mathrm{N}_T(0, \sigma^2)$, truncated to the interval
$(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$
to ensure vehicle speed is always positive and less than 30~m/s
(approximately 110~km/h).

The second vehicle behaviour requires a model of bus stop wait times.
When the vehicle is approaching stop $j$,
it stops with probability $\pi_j$.
If it does not stop, the remaining travel time $\delta_k$ is unaltered;
otherwise, the bus waits for a minimum of $\gamma$ seconds---this 
accounts for deceleration, opening and closing of doors, and acceleration
\citep{Hans_2015}---and then waits while passengers board and disembark,
which we assume to follow an exponential distribution with mean $\tau_j$.
Stopping behaviour is expressed in the following model:
\begin{equation}
\label{eq:trans_stop}
\begin{split}
p_j^{(i)} &\sim \mathrm{Bernoulli}(\pi_j) \\
\tilde d_j^{(i)} &\sim \mathrm{Exponential}(\tau_j) \\
d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}).
\end{split}
\end{equation}

The model components are implemented as an algorithm,
in which (\ref{eq:trans_newton}) is evaluated iteratively one second at a time,
each step decreasing $\delta_k^{(i)}$ by one second,
after initialising $\delta_k^{(i)} = \delta_k$.
In order to obtain the desired non-constant speed between observations,
$\epsilon_k^{(i)}$ is re-sampled each second,
allowing the particle's speed to vary.
This iterative approach allows the particle to approach the next stop,
at which point (\ref{eq:trans_stop}) is evaluated for the particle
and the dwell time subtracted from $\delta_k^{(i)}$.
This process continues until $\delta_k^{(i)}$ reaches zero.
The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
along each segment $\ell$ of the route,
storing the times so that the posterior distribution of travel time
can be computed once all particles have completed segment $\ell$:
\begin{equation}
\label{eq:post_tt}
p(z_\ell | \bY_{1:k}) \approx
    \sum_{i=1}^N W_k^{(i)} \dot\delta(z_\ell - z_\ell^{(i)}).
\end{equation}

>>>>

The particle filter provides us with a flexible method
with which to model bus behaviour.
Currently, this model includes
\begin{enumerate}
\item non-constant speed along roads (between observations), and
\item bus stop behaviour, (optional stopping and wait times while passengers board and disembark).
\end{enumerate}
The transition function consists of two models,
one for each of the above behaviours.
The first models the speed and distance travelled by the vehicle,
\begin{equation}
\label{eq:trans_newton}
\bX_k^{(i)} = f(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) = 
    \begin{bmatrix}
        x_{k-1}^{(i)} + \delta_k(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
        \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
    \end{bmatrix},
\end{equation}
where $\delta_k = t_k - t_{k-1}$
and $\epsilon_k^{(i)}\sim\mathrm{N}_T(0, \sigma^2)$, truncated to the interval
$(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$
to ensure vehicle speed is always positive and less than 30~m/s
(approximately 110~km/h).

The second vehicle behaviour requires a model of bus stop wait times.
When approaching stop $j$,
the vehicle stops with probability $\pi_j$.
If it does not stop, the remaining travel time $\delta_k$ is unaltered;
otherwise, the bus waits for a minimum of $\gamma$ seconds---% 
accounting for deceleration, opening and closing of doors, and acceleration
\citep{Hans_2015}---and then waits while passengers board and disembark,
which we assume to follow an exponential distribution with mean $\tau_j$.
Stopping behaviour is expressed in the following model:
\begin{equation}
\label{eq:trans_stop}
\begin{split}
p_j^{(i)} &\sim \mathrm{Bernoulli}(\pi_j) \\
\tilde d_j^{(i)} &\sim \mathrm{Exponential}(\tau_j) \\
d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}),
\end{split}
\end{equation}
using constant values of $\pi_j = 0.5$ and $\tau_j = 10$~seconds for all stops;
if available, historical dwell time information can easily be used
to specify each $\tau_j$,
or even replace the exponential prior altogether.
As it stands, we found these values to be vague enough that
the actual vehicle trajectory was sampled by enough of the particles
to prevent degeneration of the filter;
the consequence being increased variance of travel time estimates.

The model components are implemented as an algorithm,
in which (\ref{eq:trans_newton}) is evaluated iteratively one second at a time,
each step decreasing $\delta_k^{(i)}$ by one second,
after initialising $\delta_k^{(i)} = \delta_k$.
% To allow non-constant speed between observations,
We also re-sampled
$\epsilon_k^{(i)}$ each second
to allow the particle's speed to vary between observations.
This iterative approach allows the particle to approach the next stop,
at which point (\ref{eq:trans_stop}) is evaluated for the particle
and the dwell time subtracted from $\delta_k^{(i)}$.
This process continues until $\delta_k^{(i)}$ reaches zero.
The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
along each segment $\ell$ of the route,
storing the times so that the posterior distribution of travel time
can be computed once all particles have completed segment $\ell$:
\begin{equation}
\label{eq:post_tt}
p(z_\ell | \bY_{1:k}) \approx
    \sum_{i=1}^N W_k^{(i)} \dot\delta(z_\ell - z_\ell^{(i)}).
\end{equation}

### 2019-08-12 anzjs2018/sections/05_discuss.tex
minor edits

We have focussed on making a general framework
in which the existing models used for specific components
can easily be changed---%
for example, using a more complex model of stop dwell times
based on historical data in place of the exponential---%
or new ones added, such as implementing intersection-specific behaviour
if intersection locations are available.
Future work is planned to detect route meeting points
and use these as intersection nodes
 to further reduce road segment overlap.

>>>>

We have focussed on making a general framework
into which new models can be incorporated,
such as implementing intersection-specific behaviour
once intersection locations are available,
or exiting models changed.
For example, we plan to model stop dwell time using historical data,
which will allow us to replace the exponential distribution
with something more appropriate.
